---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.path  = "man/figures/README-",
  out.width = "100%",
  message   = FALSE,
  warning   = FALSE
)
```

# BETS

<!-- badges: start -->
[![R-CMD-check](https://github.com/LorenzoZambon/BETS/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/LorenzoZambon/BETS/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/LorenzoZambon/BETS/graph/badge.svg)](https://app.codecov.io/gh/LorenzoZambon/BETS)
<!-- badges: end -->

**BETS** implements a fully Bayesian ETS state-space model,
which accounts for both parameter and model uncertainty.
Currently, only additive error models are supported.
Initial states and the observation variance are analytically marginalized via conjugate
Normal-Inverse-Chi-Squared priors; only the smoothing parameters (α, β, γ, φ)
are sampled via Adapive Importance Sampling (AIS). 
Multiple ETS variants are combined by Bayesian Model Averaging
(BMA, default) or stacking. 
Forecasts are generated by `predict()` as sample-path trajectories.

Key features:

- **Parameter uncertainty**: parameters are sampled from the posterior distribution.
- **Model uncertainty**: model combination is implemented via BMA (default) or 
stacking over the additive-error ETS model family.
- **Fast**: Rao-Blackwellization reduces the dimension of the sampling space; 
likelihood evaluation is vectorized and implemented in C++. 


## Installation

```r
# TODO
```

## Example: M3 monthly series 

First, load the packages and set the seed.

```{r load-pkgs}
library(BETS)
library(forecast)   # for ets() comparison
library(Mcomp)      # M3 competition data
library(ggplot2)    # for plotting

set.seed(123)
```

We now fit BETS and ETS (from the forecast package) to a series from the M3 dataset.
ETS uses a maximum likelihood point estimate of the parameters, 
while BETS accounts for uncertainty over these parameters by drawing samples 
from the posterior distribution.
Moreover, `ets()` selects a single model by AICc, BETS instead automatically combines multiple ETS variants.

```{r fit}
# Pick a time series from the M3 dataset
freq_series <- "MONTHLY"  # either "MONTHLY", "QUARTERLY", or "YEARLY"
j <- 10  # index of series to fit and plot 
series <- Filter(function(s) s$period == freq_series, M3)[[j]]
y_train <- series$x                 
y_test  <- series$xx               
h       <- length(y_test)

# Fit models 
bets_fit <- BETS::bets(y_train)                               
ets_fit  <- forecast::ets(y_train, additive.only = TRUE) 

# Print details on the fitted models
print(ets_fit)    # print ETS model selected by AIC
print(bets_fit)   # print model weights
```

We then compute forecasts from both models.

```{r forecast}
# Compute forecasts
bets_fc <- predict(bets_fit, h = h, level = c(80, 95))
ets_fc  <- forecast(ets_fit, h = h, level = c(80, 95))

# Print forecast means and prediction intervals
# TODO
```

Finally, we plot the forecasts from both models, including the 80% and 95% prediction intervals, 
and compare them to the actual holdout values.

```{r plot, echo = FALSE}
# ---- Build data frames for ggplot2 --------------------------------------- #
t_train <- as.numeric(time(y_train))
t_fc    <- as.numeric(time(bets_fc$mean))
t_test  <- as.numeric(time(y_test))

df_train <- data.frame(t = t_train, y = as.numeric(y_train))
df_test  <- data.frame(t = t_test,  y = as.numeric(y_test))

df_fc <- rbind(
  data.frame(
    t     = t_fc,
    mean  = as.numeric(bets_fc$mean),
    lo80  = as.numeric(bets_fc$lower[, "80"]),
    hi80  = as.numeric(bets_fc$upper[, "80"]),
    lo95  = as.numeric(bets_fc$lower[, "95"]),
    hi95  = as.numeric(bets_fc$upper[, "95"]),
    model = "BETS"
  ),
  data.frame(
    t     = t_fc,
    mean  = as.numeric(ets_fc$mean),
    lo80  = as.numeric(ets_fc$lower[, "80%"]),
    hi80  = as.numeric(ets_fc$upper[, "80%"]),
    lo95  = as.numeric(ets_fc$lower[, "95%"]),
    hi95  = as.numeric(ets_fc$upper[, "95%"]),
    model = "ETS"
  )
)

# ---- Plot ---------------------------------------------------------------- #
ggplot() +
  geom_ribbon(
    data  = df_fc,
    aes(x = t, ymin = lo95, ymax = hi95, fill = model),
    alpha = 0.15
  ) +
  geom_ribbon(
    data  = df_fc,
    aes(x = t, ymin = lo80, ymax = hi80, fill = model),
    alpha = 0.3
  ) +
  geom_line(
    data  = df_train,
    aes(x = t, y = y),
    colour    = "grey40",
    linewidth = 0.6
  ) +
  geom_line(
    data      = df_fc,
    aes(x = t, y = mean, colour = model),
    linewidth = 0.8
  ) +
  geom_line(
    data      = df_test,
    aes(x = t, y = y),
    colour    = "black",
    linewidth = 0.8,
    linetype  = "dashed"
  ) +
  scale_colour_manual(values = c("BETS" = "#E64B35", "ETS" = "#4DBBD5")) +
  scale_fill_manual(  values = c("BETS" = "#E64B35", "ETS" = "#4DBBD5")) +
  labs(
    title    = paste0("M3 monthly series: ", series$sn),
    subtitle = paste0(h, "-step-ahead forecasts with 80% and 95% prediction intervals"),
    x        = NULL,
    y        = NULL,
    colour   = NULL,
    fill     = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

