---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.path  = "man/figures/README-",
  out.width = "100%",
  message   = FALSE,
  warning   = FALSE
)
```

# BayesETS

<!-- badges: start -->
[![R-CMD-check](https://github.com/LorenzoZambon/BETS/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/LorenzoZambon/BETS/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/LorenzoZambon/BETS/graph/badge.svg)](https://app.codecov.io/gh/LorenzoZambon/BETS)
<!-- badges: end -->

**BayesETS** implements Bayesian ETS (exponential smoothing) state-space models,
which account for both parameter and model uncertainty.
Currently, only additive models are supported.
Initial states and the observation variance are analytically marginalized via conjugate
Normal-Inverse-Chi-Squared priors.
The smoothing parameters ($\alpha, \beta, \gamma, \phi$) are either integrated by adaptive
Gauss-Hermite quadrature (AGHQ) or sampled via Adaptive Importance Sampling (AIS).
Multiple ETS models are combined by Bayesian Model Averaging (BMA).
Forecasts are generated by `predict()` as sample-path trajectories.

Key features:

- **Parameter uncertainty**: parameters are integrated over their posterior distribution,
  rather than estimated as point values.
- **Model uncertainty**: the additive ETS models are combined via BMA, rather than selecting
  a single one.
- **Fast**: analytical marginalization reduces the dimension of the integration space to
  (at most) 4; likelihood evaluation is vectorized and implemented in C++.
  Running on 1 core of a laptop with an AMD Ryzen 5, on the M3 data `bets()` requires on average 
  0.12 s per monthly series, 0.07 s per quarterly and 0.02 s per yearly one
  (median length: 115, 44 and 19 observations respectively).


## Installation

You can install the development version from [GitHub](https://github.com/LorenzoZambon/BETS):

```r
# install.packages("remotes")
remotes::install_github("LorenzoZambon/BETS")
```

The package contains C++ code: installing it from GitHub requires a compiler
(Rtools on Windows, the Xcode command line tools on macOS, a C++ compiler on Linux).

## Example: M3 monthly series

We fit `bets()` and `ets()` (from the forecast package) to a monthly series of the M3
competition, and compare their forecasts with the observed test set.

`ets()` estimates the parameters of each model via maximum likelihood and selects a single model by AICc.
`bets()` fits all the additive models and averages them, with weights given by their
posterior probabilities; within each model, it integrates over the posterior of the parameters.

```{r load-pkgs}
library(BayesETS)
library(forecast)   # for comparison with ets()
library(Mcomp)      # M3 competition data
library(ggplot2)    # for plot

set.seed(123)
```

```{r fit}
series  <- M3[["N2471"]]
y_train <- series$x
y_test  <- series$xx
h       <- length(y_test)

bets_fit <- bets(y_train, additive.only = TRUE)
ets_fit  <- ets(y_train, additive.only = TRUE)
```

**Models.** `ets()` selects a single model, while `bets()` gives a weight to each one:

```{r models}
ets_fit$method   # model selected by AICc

bets_fit         # weights of the models, posterior means of the parameters
```

**Parameters.** For the model selected by `ets()`, we compare its maximum likelihood estimates
with the posterior of the same model in `bets()` (mean, standard deviation and 95% credible
interval, from `summary()`):

```{r params}
model <- gsub("ETS|[(),]", "", ets_fit$method)    # e.g. "AAA"
post  <- summary(bets_fit)$parameters[[model]]
ml    <- c(ets_fit$par[setdiff(rownames(post), "sigma")], sigma = sqrt(ets_fit$sigma2))
round(cbind(ets = ml, post), 3)
```


Forecasts are obtained with `predict()`, by simulating future trajectories.

```{r forecast}
bets_fc <- predict(bets_fit, h = h, level = c(80, 95))
ets_fc  <- forecast(ets_fit, h = h, level = c(80, 95))

bets_fc
```

Finally, we plot the last years of the series with the forecasts of both models
(80% and 95% prediction intervals) and the observed test set (dashed).

```{r plot, echo = FALSE, fig.height = 4.5}
# ---- Data frames for ggplot2 ---------------------------------------------- #
y_show <- window(y_train, start = time(y_train)[length(y_train) - 47])   # last 4 years
t_fc   <- as.numeric(time(bets_fc$mean))

df_train <- data.frame(t = as.numeric(time(y_show)), y = as.numeric(y_show))
df_test  <- data.frame(t = as.numeric(time(y_test)), y = as.numeric(y_test))

df_fc <- rbind(
  data.frame(
    t     = t_fc,
    mean  = as.numeric(bets_fc$mean),
    lo80  = as.numeric(bets_fc$lower[, "80"]),
    hi80  = as.numeric(bets_fc$upper[, "80"]),
    lo95  = as.numeric(bets_fc$lower[, "95"]),
    hi95  = as.numeric(bets_fc$upper[, "95"]),
    model = "BETS"
  ),
  data.frame(
    t     = t_fc,
    mean  = as.numeric(ets_fc$mean),
    lo80  = as.numeric(ets_fc$lower[, "80%"]),
    hi80  = as.numeric(ets_fc$upper[, "80%"]),
    lo95  = as.numeric(ets_fc$lower[, "95%"]),
    hi95  = as.numeric(ets_fc$upper[, "95%"]),
    model = "ETS"
  )
)

# ---- Plot ----------------------------------------------------------------- #
ggplot() +
  geom_ribbon(
    data  = df_fc,
    aes(x = t, ymin = lo95, ymax = hi95, fill = model),
    alpha = 0.15
  ) +
  geom_ribbon(
    data  = df_fc,
    aes(x = t, ymin = lo80, ymax = hi80, fill = model),
    alpha = 0.3
  ) +
  geom_line(
    data      = df_train,
    aes(x = t, y = y),
    colour    = "grey40",
    linewidth = 0.6
  ) +
  geom_line(
    data      = df_fc,
    aes(x = t, y = mean, colour = model),
    linewidth = 0.8
  ) +
  geom_line(
    data      = df_test,
    aes(x = t, y = y),
    colour    = "black",
    linewidth = 0.8,
    linetype  = "dashed"
  ) +
  scale_colour_manual(values = c("BETS" = "#E64B35", "ETS" = "#4DBBD5")) +
  scale_fill_manual(  values = c("BETS" = "#E64B35", "ETS" = "#4DBBD5")) +
  labs(
    title    = paste0("M3 monthly series ", series$sn),
    subtitle = paste0(h, "-step-ahead forecasts with 80% and 95% prediction intervals"),
    x        = NULL,
    y        = NULL,
    colour   = NULL,
    fill     = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```
