---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.path  = "man/figures/README-",
  out.width = "100%",
  message   = FALSE,
  warning   = FALSE
)
```

# BayesETS

<!-- badges: start -->
[![R-CMD-check](https://github.com/LorenzoZambon/BayesETS/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/LorenzoZambon/BayesETS/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/LorenzoZambon/BayesETS/graph/badge.svg)](https://app.codecov.io/gh/LorenzoZambon/BayesETS)
<!-- badges: end -->

**BayesETS** implements Bayesian ETS (exponential smoothing) state-space models,
which account for both parameter and model uncertainty.
Currently, only additive models are supported.
Initial states and the observation variance are analytically marginalized via conjugate
Normal-Inverse-Chi-Squared priors.
The smoothing parameters ($\alpha, \beta, \gamma, \phi$) are integrated numerically,
either by adaptive Gauss-Hermite quadrature (AGHQ) or by adaptive importance sampling (AIS).
Multiple ETS models are combined by Bayesian Model Averaging (BMA).
Forecasts are generated by `predict()`, by simulating future sample paths.

Key features:

- **Parameter uncertainty**: parameters are integrated over their posterior distribution,
  rather than estimated as point values.
- **Model uncertainty**: multiple ETS models are combined via BMA, rather than selecting
  a single one.
- **Fast**: analytical marginalization reduces the dimension of the integration space to
  (at most) 4; likelihood evaluation is vectorized and implemented in C++.
  On the M3 data, `bets()` takes on average 0.12 s per monthly series, 0.06 s per quarterly
  and 0.02 s per yearly one (median lengths: 115, 44 and 19 observations), on one core of a
  laptop with an AMD Ryzen 5.


## Installation

You can install the development version from [GitHub](https://github.com/LorenzoZambon/BayesETS):

```r
# install.packages("remotes")
remotes::install_github("LorenzoZambon/BayesETS")
```

The package contains C++ code: installing it from GitHub requires a compiler
(Rtools on Windows, the Xcode command line tools on macOS, a C++ compiler on Linux).

## Example: M3 monthly series

We fit `bets()` and `ets()` (from the forecast package) to a monthly series of the M3
competition, and compare the fitted models and their forecasts.
`ets()` estimates the parameters of each model by maximum likelihood and selects a single model
by AICc. `bets()` fits all the additive models and averages them, with weights given by their
posterior probabilities; within each model, it integrates over the posterior of the parameters.

```{r load-pkgs}
library(BayesETS)
library(forecast)   # for comparison with ets()
library(Mcomp)      # M3 competition data
library(ggplot2)    # for plot

set.seed(123)
```

```{r fit}
series  <- M3[["N2542"]]
y_train <- series$x
y_test  <- series$xx
h       <- length(y_test)

bets_fit <- bets(y_train, additive.only = TRUE)
ets_fit  <- ets(y_train, additive.only = TRUE)
```

**Overview.** The print methods summarise the two fits:

```{r overview}
ets_fit    # selected model, estimates of the parameters and of the initial states

bets_fit   # model weights, posterior means of the parameters
```

**Models.** `ets()` selects a single model by AICc, while `bets()` gives a weight to each model,
its posterior probability (from `summary()`, with the log evidence of each model):

```{r models}
ets_fit$method

s <- summary(bets_fit)
s$models
```

**Parameters.** For the model selected by `ets()`, ETS(A,A,A), we compare the maximum likelihood
estimates with the posterior of the same model in `bets()` (mean, standard deviation and 95%
credible interval):

```{r params}
# ets(): maximum likelihood estimates
round(c(ets_fit$par[c("alpha", "beta", "gamma")], sigma = sqrt(ets_fit$sigma2)), 4)

# bets(): posterior of the same model
post <- s$parameters$AAA
post
```

`ets()` gives a single value for each parameter, while `bets()` gives its posterior
distribution, summarised here by mean, standard deviation and 95% credible interval.
For example, `ets()` estimates $\alpha$ = `r round(ets_fit$par[["alpha"]], 2)`, while the 95%
credible interval goes from `r round(post["alpha", "2.5%"], 2)` to
`r round(post["alpha", "97.5%"], 2)`; and `ets()` sets $\beta$ and $\gamma$ to zero (their
lower bound: constant slope and seasonal pattern), while the posterior also gives weight to
values up to about `r round(post["beta", "97.5%"], 2)` and `r round(post["gamma", "97.5%"], 2)`.
`bets()` carries this uncertainty into the forecasts.

Forecasts are obtained with `predict()`, by simulating future trajectories.

```{r forecast}
bets_fc <- predict(bets_fit, h = h, level = c(80, 95))
ets_fc  <- forecast(ets_fit, h = h, level = c(80, 95))

bets_fc
```

Finally, we plot the last years of the series with the forecasts of both models
(80% and 95% prediction intervals) and the observed test set (dashed).
The intervals of `bets()` account for the uncertainty on the parameters and on the model, while
those of `ets()` are conditional on the estimates of the selected model.

```{r plot, echo = FALSE, fig.height = 4.5}
# ---- Data frames for ggplot2 ---------------------------------------------- #
y_show <- window(y_train, start = time(y_train)[length(y_train) - 47])   # last 4 years
t_fc   <- as.numeric(time(bets_fc$mean))

df_train <- data.frame(t = as.numeric(time(y_show)), y = as.numeric(y_show))
df_test  <- data.frame(t = as.numeric(time(y_test)), y = as.numeric(y_test))

df_fc <- rbind(
  data.frame(
    t     = t_fc,
    mean  = as.numeric(bets_fc$mean),
    lo80  = as.numeric(bets_fc$lower[, "80"]),
    hi80  = as.numeric(bets_fc$upper[, "80"]),
    lo95  = as.numeric(bets_fc$lower[, "95"]),
    hi95  = as.numeric(bets_fc$upper[, "95"]),
    model = "BETS"
  ),
  data.frame(
    t     = t_fc,
    mean  = as.numeric(ets_fc$mean),
    lo80  = as.numeric(ets_fc$lower[, "80%"]),
    hi80  = as.numeric(ets_fc$upper[, "80%"]),
    lo95  = as.numeric(ets_fc$lower[, "95%"]),
    hi95  = as.numeric(ets_fc$upper[, "95%"]),
    model = "ETS"
  )
)

# ---- Plot ----------------------------------------------------------------- #
ggplot() +
  geom_ribbon(
    data  = df_fc,
    aes(x = t, ymin = lo95, ymax = hi95, fill = model),
    alpha = 0.15
  ) +
  geom_ribbon(
    data  = df_fc,
    aes(x = t, ymin = lo80, ymax = hi80, fill = model),
    alpha = 0.3
  ) +
  geom_line(
    data      = df_train,
    aes(x = t, y = y),
    colour    = "grey40",
    linewidth = 0.6
  ) +
  geom_line(
    data      = df_fc,
    aes(x = t, y = mean, colour = model),
    linewidth = 0.8
  ) +
  geom_line(
    data      = df_test,
    aes(x = t, y = y),
    colour    = "black",
    linewidth = 0.8,
    linetype  = "dashed"
  ) +
  scale_colour_manual(values = c("BETS" = "#E64B35", "ETS" = "#4DBBD5")) +
  scale_fill_manual(  values = c("BETS" = "#E64B35", "ETS" = "#4DBBD5")) +
  labs(
    title    = paste0("M3 monthly series ", series$sn),
    subtitle = paste0(h, "-step-ahead forecasts with 80% and 95% prediction intervals"),
    x        = NULL,
    y        = NULL,
    colour   = NULL,
    fill     = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

## Benchmark

We compare `bets()` and `ets()` (additive models) on all the yearly, quarterly and monthly
series of the M3 data (2829 series), with the M3 test sets (horizons 6, 8 and 18):

- **MASE**: mean absolute error of the point forecasts, scaled by the in-sample mean absolute
  error of the seasonal naive forecast;
- **Coverage** of the 80% and 95% prediction intervals;
- **MSIS**: interval score of the 95% intervals, with the same scaling (lower is better);
- **Time**: seconds per series, for fit and forecast.

```{r benchmark, echo = FALSE, results = "asis"}
# HTML table (GitHub markdown has no multirow): one cell per period, spanning ets and bets
res <- read.csv("benchmark/m3_results.csv")
groups <- c(YEARLY = "Yearly", QUARTERLY = "Quarterly", MONTHLY = "Monthly")
cells <- function(...) paste0("<td align=\"right\">", c(...), "</td>", collapse = "")
rows <- character()
for (g in names(groups)) {
  d <- res[res$period == g, ]
  for (m in c("ets", "bets")) {
    s <- colMeans(d[d$method == m, c("MASE", "cov80", "cov95", "MSIS", "time")])
    period <- if (m == "ets") {
      sprintf("<td rowspan=\"2\">%s (%d)</td>", groups[[g]], length(unique(d$series)))
    }
    rows <- c(rows, paste0("<tr>", period, "<td>", m, "</td>",
                           cells(sprintf("%.3f", s[1:3]), sprintf("%.2f", s[4]), sprintf("%.3f", s[5])),
                           "</tr>"))
  }
}
# Raw HTML block, passed to README.md as it is
cat("```{=html}", "<table>",
    "<thead><tr><th>Period (series)</th><th>Method</th><th>MASE</th><th>Coverage 80%</th>",
    "<th>Coverage 95%</th><th>MSIS</th><th>Time (s)</th></tr></thead>",
    "<tbody>", rows, "</tbody>", "</table>", "```", sep = "\n")
```

The script is in [`benchmark/m3_benchmark.R`](benchmark/m3_benchmark.R).

## Limitations

- Only additive models are supported (additive error, trend and seasonality).
- The series must not contain missing values.
- Exogenous regressors are not supported.
